Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-08-12 tex/manuscript.tex
Noen siste små justeringer

\footnotetext{\color{blue}Manuscript version \today.}

@@

% List of authors, with corresponding author marked by asterisk
\author{VEGARD NYGAARD, EINAR ANDREAS RØDLAND\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[8pt]
EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
\textit{Institute of Cancer Genetics and Informatics, Oslo University Hospital HF - Radiumhospitalet,
Montebello, 
0310 Oslo,
Norway}
\\[2pt]
\textit{Department of Informatics, University of Oslo, 
0316 OSLO
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@ifi.uio.no}}

@@

Two popular tools for batch adjustment of gene expression data are the R bioconductor packages \texttt{limma} (\citealp{Smyth2003}) and ComBat (\citealp{Johnson2007}) which is now included in the \texttt{sva} package (\citealp{Leek2012}). Both these packages allow the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, ComBat uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora, and ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported.

When sample or batch sizes are small, statisticians would most likely take extra precautions. However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful on large samples and batch sizes as for small. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

@@

The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner. Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies. The batch effect estimation errors will influence group effects in proportion to how well the group is represented in each batch:
\begin{equation}
\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}
=\frac{1}{n_{\textrm{-}j}} \sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}
=\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i)
\quad\text{where}\quad
n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}
\end{equation}
so that
\begin{equation}
\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
=\beta_j-\beta_{j'}+\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'}
 -\sum_{i=1}^m \left(\frac{n_{ij}}{n_{\textrm{-}j}}-\frac{n_{ij'}}{n_{\textrm{-}j'}}\right)(\hat{\gamma}_i-\gamma_i).
\end{equation}
In a balanced group--batch design, the estimation error $\hat\gamma_i-\gamma_i$ has the same effect for all groups, and thus does not influence group comparisons. In an unbalanced design, however, it will induce systematic differences between groups which, when ignored in downstream analyses, may lead to biases or over-confidence in estimated group differences. In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data. For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library has a executable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

@@

The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance. We have reanalysed to cases with varying degree of unbalance.

@@

Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat. Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat. However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done. In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \citealp{Chen2011}). A brief inquiry into  some of the articles citing ComBat (574 on Google Scholar) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences.

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb ((\citealp{Taminau2011}), presently without covariates as an option.

@@

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
In section~\label{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computations, although more complicated, can be made for multiple groups. However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.

@@

[...Acknowledgements...]
{\it Conflict of Interest}: None declared.

>>>>

\renewcommand{\thefootnote}{§}
\footnotetext{\color{blue}Manuscript version \today.}

@@

% List of authors, with corresponding author marked by asterisk
\renewcommand{\thefootnote}{\textdagger}
\footnotetext{These two authors contributed equally.}
\author{VEGARD NYGAARD$^\dagger$, EINAR ANDREAS RØDLAND$^\dagger$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[8pt]
EIVIND HOVIG$^\ast$\\[4pt]
% Author addresses
\textit{Department of Tumor Biology,
Institute for Cancer Research,
Oslo University Hospital HF - Radiumhospitalet,
Montebello,
0310 Oslo,
Norway}
\\[2pt]
\textit{Institute of Cancer Genetics and Informatics, Oslo University Hospital HF - Radiumhospitalet,
Montebello, 
0310 Oslo,
Norway}
\\[2pt]
\textit{Department of Informatics, University of Oslo, 
0316 OSLO
Norway}
\\[2pt]
% E-mail address for correspondence
{ehovig@ifi.uio.no}}

@@

A popular tool for batch adjustment of gene expression data is ComBat (\citealp{Johnson2007}) which is now included in the R bioconductor package \texttt{sva} (\citealp{Leek2012}). This allows the inclusion of covariates, e.g. group difference, the effects of which should not be removed in the batch adjustment. In addition, it uses an empirical Bayes approach to avoid over-correction for batch effects, which is critical for small batches. Increasingly, the inclusion of group difference as a covariate when removing batch effects has been recommended, both in the \texttt{sva} tutorial and user fora. ComBat has thus helped popularise this approach. Based on the actual use of ComBat by the authors and others, ourselves included, we suspect that adjusted data are commonly treated as ``batch effect free'' in subsequent analyses, and as a consequence, confidence in group effects has been overestimated and false results reported. The R package \texttt{limma} (\citealp{Smyth2003}) also offers batch adjustment with covariates, but notes that this is not intended used prior to linear modeling.

When sample or batch sizes are small, statisticians would most likely take extra precautions. However, batch adjustment using two-way ANOVA or ComBat on unbalanced data sets may be just as harmful for large samples and batch sizes as for small. For example, group comparisons using one-way ANOVA on the batch adjusted data will essentially result in $F$ statistics that are inflated by a fixed factor which depends on the unevenness of the design, rather than the size of the sample or batches. The effect of this may be further exacerbated by running these analyses a large number of times, e.g. on thousands of genes, and use false discovery rate to determine significant cases: an approach that is particularly sensitive to inflated false positive rates.

@@

The estimation error $\hat\gamma_i-\gamma_i$ affects all values within the same batch in the same manner. Thus, while the aim is to remove spurious dependencies within batches, it may also induce new dependencies. The batch effect estimation errors will influence group effects in proportion to how well the group is represented in each batch:
\begin{equation}
\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}
=\frac{1}{n_{\textrm{-}j}} \sum_{i=1}^m\sum_{r=1}^{n_{ij}}\widetilde{Y}_{ijr}
=\alpha+\beta_j+\bar{\epsilon}_{\textrm{-}j}
  -\sum_{i=1}^m \frac{n_{ij}}{n_{\textrm{-}j}}(\hat{\gamma}_i-\gamma_i)
\quad\text{where}\quad
n_{\textrm{-}j}=\sum_{i=1}^m n_{ij}
\end{equation}
so that
\begin{equation}
\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j}-\widetilde{Y}^{\textrm{cov}}_{\textrm{-}j'}
=(\beta_j-\beta_{j'})+(\bar{\epsilon}_{\textrm{-}j}-\bar{\epsilon}_{\textrm{-}j'})
 -\sum_{i=1}^m \left(\frac{n_{ij}}{n_{\textrm{-}j}}-\frac{n_{ij'}}{n_{\textrm{-}j'}}\right)(\hat{\gamma}_i-\gamma_i).
\end{equation}
In a balanced group--batch design, the estimation error $\hat\gamma_i-\gamma_i$ has the same effect for all groups, and thus does not influence group comparisons. In an unbalanced design, however, it will induce systematic differences between groups which, when ignored in downstream analyses, may lead to biases or over-confidence in estimated group differences. In Figure~\ref{fig:boxplots}d, the effect of batch correction using group as covariate is shown with confidence intervals of the corrected data. For comparison, least square means estimates (R package \texttt{lsmeans}) are used in Figure~\ref{fig:boxplots}e to illustrate more appropriate confidence intervals that incorporate the uncertainties of the estimation errors.

@@

The undesired consequences of preserving group effects when correcting for batch effect is readily illustrated with a sanity check using random numbers. The documentation accompanying the \texttt{sva} library has an executable example demonstrating how to adjust a data set with ComBat, followed by an F-test. Replacing the real data with random numbers from a standard normal distribution (mean=0, sd=1), but otherwise following the instructions, will generate the p-value distribution shown in Figure~\ref{fig:sanity}a.

@@

The extent to which batch adjustment with group differences retained will confound subsequent analyses depends on the batch--group balance. We have reanalysed two cases with varying degree of unbalance.

@@

Our knowledge of the problem discussed in this article came through a typical use case when trying to adjust for batch effects in an unbalanced data set using ComBat. Upon realizing that the confidence in the estimated group differences was exaggerated, we searched the literature for a better understanding of correct use and the potential limitations of ComBat. However, the authors of ComBat and the \texttt{sva} package recommended including study group as a covariate as we had done. In addition, other studies investigating batch effects were mostly recommending ComBat without much concern (\citealp{Kupfer2012}, \citealp{Kitchen2011}, \citealp{Chen2011}). A brief inquiry into some of the articles citing ComBat (466 in Web of Science) revealed few reported problems, although their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences. Some used ComBat with covariates, some without, but we did not find any that addressed the potential problems related to use of covariates.

A further indication of the carefree use of the procedure was the frequent omittance of program parameters that were applied, i.e batch labels or whether group labels were supplied as covariates. Often, no effort was undertaken in order to substantiate the existence of batch effects, beyond stating the presence of batches. The incorporation of ComBat into various analysis tools appears to make it more accessible, but its use less transparent. Such tools include GenePattern (\citealp{Reich2006}), which offers ComBat with covariates, as well as TCGA and  inSilicoDb (\citealp{Taminau2011}), presently without covariates as an option.

@@

A superficial assessment of the reliability of results based on batch corrected data with group differences retained can often be made.
In section~\ref{sec:twogroups}, the ratio $\nu_0/\nu\ge1$ indicates by how much the effective sample size is overestimated when comparing two groups. A similar computation, although more complicated, can be made for multiple groups. However, the two-group assessment should still give a fair idea of the reliability of pairwise group comparisons. A major concern would be batches where either of the groups of interest are missing or strongly under-represented, which would contribute to the nominal sample size, but not to the effective sample size.

@@

We are grateful to Geir Kjetil Sandve for useful suggestions and discussions.

This project was in part financed through the EUROCAN platform (VN) and the MetAction project (EAR).

{\it Conflict of Interest}: None declared.
